CdGF_aucrf

### 2016-09-20 code/CdGF_AUCRF.Rmd
add cfu invsimp and entero to AUCRF

meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

@@

Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'Euth_Early')], 
                               rel_abund_stool, by.x = 'mouse_id', by.y='row.names')
Predict_early_euth_df_names <- Predict_early_euth_df$mouse_id
Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)

@@

#### 10-fold Cross validated - 10 repetitions
start <- proc.time()[3]
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp <- c()
cv10f_all_pred <- c()
for(j in 1:iters){
  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
  cv10f_probs <- rep(NA,nrow(aucrf_data))
  for(i in seq(1,nrow(aucrf_data),5)){
    train <- aucrf_data[sampling[-(i:(i+4))],]
    test <- aucrf_data[sampling[i:(i+4)],]
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
  cv10f_aucs[j] <- cv10f_roc$auc
}
cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
end <- proc.time()[3]
time <- end-start

@@

#### leave-one-cage-out
LOcO_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'cage_id', 'Euth_Early')], 
                               rel_abund_stool[ , rf_early_aucrf$Xopt], by.x = 'mouse_id', by.y='row.names')
cage_list <- unique(LOcO_df$cage_id)

LOcO_probs <- NULL
LOcO_probs <- sapply(1:length(cage_list), function(i){
  by_cage <- LOcO_df$cage_id == cage_list[i]
  LOcO_predict <- select(LOcO_df, -mouse_id, -cage_id)
  train <- LOcO_predict[-by_cage,]
  test <- LOcO_predict[by_cage,]
  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
  LOcO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
})
LOcO_probs <- unlist(LOcO_probs)
LOcO_roc <- roc(aucrf_data$Euth_Early~LOcO_probs)

#### Random - 10-fold Cross validated - 10 repetitions
ran_aucrf_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early, 
                                                     nrow(Predict_early_euth_df))),
                             Predict_early_euth_df[,rf_early_aucrf$Xopt])
ran_cv10f_aucs <- c()
ran_cv10f_all_resp <- c()
ran_cv10f_all_pred <- c()
for(j in 1:iters){
  sampling <- sample(1:nrow(ran_aucrf_data),nrow(ran_aucrf_data),replace=F)
  ran_cv10f_probs <- rep(NA,nrow(ran_aucrf_data))
  for(i in seq(1,nrow(ran_aucrf_data),5)){
    train <- ran_aucrf_data[sampling[-(i:(i+4))],]
    test <- ran_aucrf_data[sampling[i:(i+4)],]
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    ran_cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  ran_cv10f_roc <- roc(ran_aucrf_data$Euth_Early~ran_cv10f_probs)
  ran_cv10f_all_pred <- c(ran_cv10f_all_pred, ran_cv10f_probs)
  ran_cv10f_all_resp <- c(ran_cv10f_all_resp, ran_aucrf_data$Euth_Early)
  ran_cv10f_aucs[j] <- ran_cv10f_roc$auc
}
ran_cv10f_roc <- roc(ran_cv10f_all_resp~ran_cv10f_all_pred)

set.seed(seed)

@@

@@

### Validate AUCRF with Randomizing

```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}
#create dataframe for RF-classification
Euth_Early <- factor(sample(c(1,0), size = nrow(rel_abund_stool), replace = T))
Validate_early_euth_df <- cbind(Euth_Early, rel_abund_stool)

@@

#### 10-fold Cross validated - 10 repetitions
start <- proc.time()[3]
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp <- c()
cv10f_all_pred <- c()
for(j in 1:iters){
  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
  cv10f_probs <- rep(NA,nrow(aucrf_data))
  for(i in seq(1,nrow(aucrf_data),5)){
    train <- aucrf_data[sampling[-(i:(i+4))],]
    test <- aucrf_data[sampling[i:(i+4)],]
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
  cv10f_aucs[j] <- cv10f_roc$auc
}
cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
end <- proc.time()[3]
time <- end-start

@@

#### leave-one-cage-out
LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id']),
                 cbind(Euth_Early, rel_abund_stool[ , rf_early_aucrf$Xopt]))
cage_list <- unique(LOcO_df$cage_id)

#### plot rocs
par(mar=c(4,4,1,1))
plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
plot(LOcO_roc, col='red', lwd=2, add=T, lty=2)
mtext(side=2, text="Sensitivity", line=2.5)
mtext(side=1, text="Specificity", line=2.5)
legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
),lty=c(1,2,1,2,0,0), lwd=2, col=c('black','blue','gray','red'), bty='n')
```

>>>>

meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
dist_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.0.03.subsample.groups.summary'
entero_file <- '~/Desktop/dmm/human_CdGF.0.03.subsample.shared.for.0.03.dmm.2.mix.posterior'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
dist_file <- read.table(dist_file, sep = '\t', header = T, row.names = 'group')
entero_file <- read.table(entero_file, sep = '\t', header = T, row.names = 'X')
source(tax_function)

@@

rownames(entero_file) <- gsub('_','-', rownames(entero_file))
dist_file <- dist_file[grep('D0', rownames(dist_file)), ]
rownames(dist_file) <- gsub('D0', '-D0', rownames(dist_file))
rownames(dist_file) <- gsub('--', '-', rownames(dist_file))

Predict_df <- 
  merge(
    merge(
      merge(
        meta_file[meta_file$day %in% 2, c('mouse_id', 'log_cfu')],
        merge(meta_file[meta_file$day %in% 0, c('cage_id', 'mouse_id', 'Euth_Early')],
              entero_file, by = 'row.names'), by = 'mouse_id'), 
      dist_file, by.x = 'Row.names', by.y = 'row.names'),
    rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')

Predict_early_euth_df <- select(Predict_df, -cage_id, 
                                -mouse_id, -Row.names, -Partition_2,
                                -label, -X, -invsimpson_lci, -invsimpson_hci)

@@

#### 10-fold Cross validated - 10 repetitions
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp <- c()
cv10f_all_pred <- c()
for(j in 1:iters){
  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
  cv10f_probs <- rep(NA,nrow(aucrf_data))
  for(i in seq(1, 40,4)){
    train <- aucrf_data[sampling[-(i:(i+4))],]
    test <- aucrf_data[sampling[i:(i+4)],]
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
  cv10f_aucs[j] <- cv10f_roc$auc
}
cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)

@@

#### leave-one-cage-out
LOcO_df <- Predict_df %>% 
                select(Euth_Early, cage_id, one_of(rf_early_aucrf$Xopt))
cage_list <- as.character(unique(LOcO_df$cage_id))

#### Random - 10-fold Cross validated - 10 repetitions
ran_aucrf_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early, 
                                                     nrow(Predict_early_euth_df))),
                             Predict_early_euth_df[,rf_early_aucrf$Xopt])
ran_cv10f_aucs <- c()
ran_cv10f_all_resp <- c()
ran_cv10f_all_pred <- c()
for(j in 1:iters){
  sampling <- sample(1:nrow(ran_aucrf_data),nrow(ran_aucrf_data),replace=F)
  ran_cv10f_probs <- rep(NA,nrow(ran_aucrf_data))
  for(i in seq(1,40,4)){
    train <- ran_aucrf_data[sampling[-(i:(i+4))],]
    test <- ran_aucrf_data[sampling[i:(i+4)],]
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    ran_cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  ran_cv10f_roc <- roc(ran_aucrf_data$Euth_Early~ran_cv10f_probs)
  ran_cv10f_all_pred <- c(ran_cv10f_all_pred, ran_cv10f_probs)
  ran_cv10f_all_resp <- c(ran_cv10f_all_resp, ran_aucrf_data$Euth_Early)
  ran_cv10f_aucs[j] <- ran_cv10f_roc$auc
}
ran_cv10f_roc <- roc(ran_cv10f_all_resp~ran_cv10f_all_pred)

@@

Would it help to add community information such as diversity or enterotype?  
Would adding initial increase/slope of CFU or day 1 CFU help predict?
But how to show since model is already approaching 1?  
* Rerun AUCRF with added features and see if they replace any OTUs

@@

### Validate AUCRF with Randomized Input

```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}
#create dataframe for RF-classification
Validate_early_euth_df <- Predict_early_euth_df
Validate_early_euth_df$Euth_Early <- sample(Validate_early_euth_df$Euth_Early)

@@

#### 10-fold Cross validated - 10 repetitions
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp <- c()
cv10f_all_pred <- c()
for(j in 1:iters){
  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
  cv10f_probs <- rep(NA,nrow(aucrf_data))
  for(i in seq(1,40,4)){
    train <- aucrf_data[sampling[-(i:(i+4))],]
    test <- aucrf_data[sampling[i:(i+4)],]
    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
  }
  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
  cv10f_aucs[j] <- cv10f_roc$auc
}
cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)

@@

#### leave-one-cage-out
LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id'],
                                  nrow(Validate_early_euth_df)),
                 Validate_early_euth_df[ , c('Euth_Early', rf_early_aucrf$Xopt)])
cage_list <- as.character(unique(LOcO_df$cage_id))

LOcO_probs <- NULL
LOcO_probs <- sapply(1:length(cage_list), function(i){
  by_cage <- LOcO_df$cage_id == cage_list[i]
  LOcO_predict <- select(LOcO_df, -cage_id)
  train <- LOcO_predict[-by_cage,]
  test <- LOcO_predict[by_cage,]
  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
  LOcO_probs[[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
})
LOcO_probs <- unlist(LOcO_probs)
LOcO_roc <- roc(aucrf_data$Euth_Early~LOcO_probs)

#### plot rocs
par(mar=c(4,4,1,1))
plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
plot(LOcO_roc, col='red', lwd=2, add=T, lty=2)
mtext(side=2, text="Sensitivity", line=2.5)
mtext(side=1, text="Specificity", line=2.5)
legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','red','gray'), bty='n')
```
